Proposal: Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-02-28 proposal/proposal.tex
abstract, related work, timeline, contributions

% \begin{abstract}
% \end{abstract}

@@

@@

@@

>>>>

\begin{abstract}
    We propose to study a previous set of algorithms that we developed for
    building mosaics, implemented in MATLAB\@. In particular, we are interested
    in the formalization of the algorithms and the relevant MATLAB structures in
    Coq and the termination of these algorithms under appropriate to-be-determined
    conditions. We present details on the algorithms along with our motivations
    for studying them. We then discuss an initial foray into related work.
    Next, we propose a timeline for our progress. Lastly, we briefly identify
    the major contributions of each author to this proposal.
\end{abstract}

@@

Because we are proving something about code that we have written ourselves,
there is not a vast body of literature specific to our algorithms from which we
can draw. There are however a number of related issues exposed by our code,
which we will discuss here. In particular, the following are of interest and
have been written about:

\begin{itemize}
    \item The correctness of MATLAB's \textit{cell array} data structure and its
        representation in Coq;
    \item The correctness of MATLAB's built-in numerical functions and matrix
        operations; and
    \item Proving termination of a particular program based on the inputs of the
        program.
\end{itemize}

Since our program deals with a large amount of pixel information in memory as a
matrix, it may be instructive to ensure that the representation of our data is
robust and that assumptions made about them lead to the results that we expect.
We are not the first group to demonstrate interest in proving properties about
matrix representations. In the early days of Coq, a
paper~\cite{magaud:hal-00955444} was published discussing the implementation of
square matrices in Coq, as well as proofs involving basic operations such as
addition and multiplication. This paper references \textit{Linear
Algebra}~\cite{lin_alg}, a Coq library written by a colleague which formalizes
a well-known linear algebra text~\cite{linear_algebra} and supports matrices with
a differing number of rows and columns. It is not known at this time how these
implementations compare with built-in Coq modules, but it will be useful to
reference external matrix proofs and see how we might make similar arguments
about matrix representations as they apply to our code.

MATLAB is a closed-source piece of software, and little has been published about
its correctness. In the official MATLAB forum, users have demonstrated interest
in verifying the correctness of MATLAB algorithms~\cite{duenisch_2013} and the
MATLAB staff reassure users that the algorithms have gone through rigorous
testing, but this assurance does not carry the same weight as a formal
verification. Indeed, as we have learned, these techniques fall on separate
parts of a correctness spectrum, with testing demonstrating correctness less
reliably than formal methods. The lack of related work in the MATLAB domain
serves as further motivation to obtain results about the correctness of a subset
of MATLAB features used in our code.

Program termination is a topic of wide interest in the computing community, as
it has both practical and of theoretical relevance given the famous halting
problem. \citeauthor{Cook_2011}~\cite{Cook_2011} argue in their article that
computer scientists are often hesitant to make claims about the termination of a
particular program, but that it is often achievable. Their work draws from the
field of compilers and makes the argument that one can prove the termination of
a program given certain inputs, provided that there is sufficient analysis of
the relation between the inputs and the possible paths that may be taken during
execution\footnote{In many contexts, this involves things like symbolic
execution (cf. Rosette~\cite{Rosette}).}. Many of the ideas in this work are
formalized by \citeauthor{BLANQUI_2011} in the \textit{CoLoR} library for Coq,
which is used primarily for proofs involving rewriting code and program
termination~\cite{BLANQUI_2011}. These efforts demonstrate that the field is
rich with material that relates termination and inputs, and, while we likely
won't make significant contributions to this work, it will be interesting to see
how well these arguments translate to the domain of scientific computing.

@@

Given the exploratory nature of this project, we may need to modify the
properties we decide to prove and the scope of the project as we learn more
about implementing our structures in Coq and decide the extent to which we can
model elements of MATLAB code. Taking this into consideration, we estimate the
following to be to be a reasonable timeline of project completion:

\begin{itemize}
    \item \textbf{03/01/21}: Project Proposal Due
    \item \textbf{03/15/21}: Represent relevant data structures in Coq, as well
        as primitive operations for matrix manipulation and accesses.
    \item \textbf{03/29/21}: Scope of project finalized with formal problem
        specification; sketch of theorems and lemmas that we expect to write in
        the process of writing our proof.
    \item \textbf{03/31/21}: Progress Report Due
    \item \textbf{04/09/21}: Significant progress made on most theorems and
        lemmas.
    \item \textbf{04/19/21}: Most theorems and lemmas should be complete; change
        scope of project where necessary.
    \item \textbf{04/24/21}: All logic should be implemented; edits for readability should be the only changes henceforth.
    \item \textbf{04/26/21}: Final Proof Due
    \item \textbf{05/03/21}: Final Paper Due
    \item \textbf{05/05/21}: Final Presentation Delivered
\end{itemize}

@@

D. B. Knoble contributed the abstract, introduction, and motivation. He also
contributed to the explanations of the algorithms. J. Do contributed the
majority of the algorithm details, as well as the related work and timeline.
